---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 4 - Due date 02/20/23"
author: "Angela Zeng"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package
library(readxl)
library(readxl)
library(ggplot2)
library(forecast)
library(tseries)
library(Kendall)
library(dplyr)
library(lubridate)
library(tidyverse)
```

## Questions 

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production". 


```{r}
#Importing data set - using xlsx package
energy_data<- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE)

read_col_names <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(energy_data) <- read_col_names
head(energy_data,5)

#transform Date column 
energy_data$Month<- as.Date(energy_data$Month)
colnames(energy_data)[1]<- "Date"
head(energy_data)

#extract columns 
colnames(energy_data)
data<- energy_data[,c(1,5)]
head(data)
tail(data) #no zeros 

#make time series data 
ts_data<- ts(data[,2], frequency = 12, start = c(1973,1))
```


## Stochastic Trend and Stationarity Tests

### Q1

Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments:
* *x* vector containing values to be differenced;
* *lag* integer indicating with lag to use;
* *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series Do the series still seem to have trend?

```{r}
#difference the time series
diff_renewableP<- diff(ts_data, lag=1, differences=1)

#create data frame for diffeerenced series 
diff_df<- data.frame(data$Date[2:597], diff_renewableP)
colnames(diff_df)<- c('Date','diff_renewableP')

#plot the differenced series 
ggplot(diff_df, aes(x=Date, y=diff_renewableP)) +
          geom_line(color="blue") +
          labs(title = 'Differenced Total Renewable Energy Production Over Time',
               y='Differenced Total Renewable Energy Production',
               x='Year')
```
The differenced series does not seem to have a trend. The data seems to move around zero. However, the spread of the differenced series grows larger over time. 

### Q2

Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in A3 using linear regression. (Hint: Just copy and paste part of your code for A3)

Copy and paste part of your code for A3 where you compute regression for Total Energy Production and the detrended Total Energy Production

```{r}
#fit linear model over Total Renewable Energy Production 
nobs <- nrow(data) 
t <- c(1:nobs) #Create vector t
linear_trend_model=lm(`Total Renewable Energy Production`~t, data=data)  
print(summary(linear_trend_model))
beta0=as.numeric(linear_trend_model$coefficients[1])  #first coefficient is the intercept term or beta0
beta1=as.numeric(linear_trend_model$coefficients[2])  #second coefficient is the slope or beta1

#Plot Total Renewable Energy Production with trend line 
print(ggplot(data, aes(x=Date, y=`Total Renewable Energy Production`)) +
        geom_line(color="blue") +
        labs(title = 'Total Renewable Energy Production Over Time',
               y='Total Renewable Energy Production (Trillion Btu)',
               x='Year')+
        geom_smooth(color="red",method="lm"))

#detrend Total Renewable Energy Production 
detrend <- data$`Total Renewable Energy Production`-(beta0+beta1*t)

#plot detrended series
ggplot(data, aes(x= Date, y= `Total Renewable Energy Production`)) +
  geom_line(color= "blue") +
  geom_line(aes(y= detrend), col="green")+
  labs(title = 'Total Renewable Energy Production Over Time (in blue) and Detrended (in green)',
               y='Total Renewable Energy Production (Trillion Btu)',
               x='Year')
```

### Q3
Create a data frame with 4 columns: month, original series, detrended by Regression Series and differenced series. Make sure you properly name all columns. Also note that the differenced series will have only 584 rows because you loose the first observation when differencing. Therefore, you need to remove the first observations for the original series and the detrended by regression series to build the new data frame.

```{r}
#Data frame - remember to not include January 1973
alldf<- data.frame(diff_df, data$`Total Renewable Energy Production`[2:597], detrend[2:597])
alldf<- alldf[,c(1,3,4,2)]
colnames(alldf)<- c('Date','Original','Detrended','Differenced')
```

### Q4
Using ggplot() create a line plot that shows the three series together. Make sure you add a legend to the plot.

```{r}
#reeorg dataframe for legend making 
alldf2 <- alldf %>%
  select(Date, Original, Detrended, Differenced) %>%
  gather(key = "Series", value = "value", -Date)
head(alldf2)

#ggplot 
ggplot(alldf2, aes(x = Date, y = value)) + 
  geom_line(aes(color = Series))+
  labs(title = 'Total Renewable Energy Production Over Time',
               y='Total Renewable Energy Production (Trillion Btu)',
               x='Year')
```


### Q5
Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the Acf() function to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?

```{r}
#lag.max: the maximum lag at which to calculate
#the acf.plot=TRUE: the acf will be plotted

#convert dataframe with all three series into a time series 
ts_alldf<- ts(alldf[,(2:4)], frequency = 12, start = c(1973,2))

#Compare ACFs
par(mfrow=c(2,2))
acf(ts_alldf[,1],lag.max=40, plot=TRUE, main=colnames(ts_alldf)[1], ylim=c(-0.5,1))
acf(ts_alldf[,2],lag.max=40, plot=TRUE, main=colnames(ts_alldf)[2], ylim=c(-0.5,1))
acf(ts_alldf[,3],lag.max=40, plot=TRUE, main=colnames(ts_alldf)[3], ylim=c(-0.5,1))
```
The differencing method was more efficient in eliminating the trend. Most but not all autocorrelation has been removed in the Differenced series. 


### Q6
Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both test. Whats the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.

```{r}
#Seasonal Mann-Kendall
SMKtest <- SeasonalMannKendall(ts_alldf[,1])
print("Results for Seasonal Mann Kendall")
print(summary(SMKtest))

#agumented dickey fuller test (ADF)
#Null hypothesis is that data has a unit root
print("Results for ADF test")
print(adf.test(ts_alldf[,1],alternative = "stationary"))
```
According to the Seasonal Mann-Kendall test, there is no seasonality in the Total Renewable Energy Production series. According to the ACF test, with a p-value of 0.9073, which we fail to reject the null hypothesis, there is a stochastic trend in the Total Renewable Energy Production series. 


### Q7

Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. 

```{r}
#use only complete years in the data 
data_cyr<- data %>%
  filter(year(Date) != 2022)

#convert to time series 
ts_data_cyr<- ts(data_cyr[,2], frequency = 12, start = c(1973,1))

#Group data in yearly steps instances
year_matrix <- matrix(ts_data_cyr,byrow=FALSE,nrow=12)
data_yearly <- colMeans(year_matrix)
my_year <- c(year(first(data_cyr$Date)):year(last(data_cyr$Date)))
data_new_yearly <- data.frame(my_year, data_yearly)

#plot to see what yearly data looks like 
ggplot(data_new_yearly, aes(x=my_year, y=data_yearly)) +
  geom_line(color="blue") +
  geom_smooth(color="red",method="lm")+
  labs(title = 'Yearly Average Total Renewable Energy Production Over Time',
               y='Total Renewable Energy Production (Trillion Btu)',
               x='Year')
```

### Q8

Apply the Mann Kendal, Spearman correlation rank test and ADF. Are the results from the test in agreement with the test results for the non-aggregated series, i.e., results for Q6?

```{r}
#Use yearly date to run Mann Kendall
print("Results of Mann Kendall on average yearly series")
print(summary(MannKendall(data_yearly)))

#Deterministic trend with Spearman Correlation Test
#doesnt work on seasonal data, only on yearly data 
print("Results from Spearman Correlation")
sp_rho=cor.test(data_yearly,my_year,method="spearman") #with cor.test you can get test statistics
print(sp_rho)

#ADF
print("Results for ADF test on yearly data/n")
print(adf.test(data_yearly, alternative = "stationary"))
```
The Mann Kendall test indicates that there is a trend present in the data. The Spearman Correlation Test indicates that series is not stationary. The ADF tests indicates that the time series has a stochastic trend. Thus, the results from the yearly data are in agreement with the test results from the non-aggregated series.
