---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Angela Zeng"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(lubridate)
library(ggplot2)

#import data set
energy_data <- read.csv(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv", header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE) 

#transform Date column 
energy_data$Month<- ym(energy_data$Month)
colnames(energy_data)[1]<- "Date"
head(energy_data)

#extract columns 
colnames(energy_data)
data<- energy_data[,c(1,4,5,6)]
head(data)
tail(data) #no zeros 
```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code from A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r}
#make time series data 
ts_data<- ts(data[,(2:4)], frequency = 12, start = c(1973,1))
name<- colnames(data)

#plot time series, Acf, and Pacf
for(i in 1:(ncol(data)-1)){
  par((mfrow=c(1,3))) #this does not work with ggplot 
  print(ggplot(data, aes(x=Date, y=data[,(1+i)])) +
          geom_line(color="blue") +
          geom_hline(yintercept = mean(data[,(1+i)]), color="red") + 
          ylab(paste0(colnames(data)[1+i], " (Trillion Btu)", sep = " "))+
          xlab("Year")+
          labs(title = paste0(colnames(data)[1+i], " Over Time", sep = " "))
  )
  Acf(ts_data[,i], lag.max=40, main=name[i+1], ylim=c(-1,1))
  Pacf(ts_data[,i], lag.max=40,main=name[i+1], ylim=c(-1,1))
}

```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Total Biomass Energy Production and Total Renewable Energy Production have an increading linear trend. Hydroelectric Power Consumption has either a seasonal or a cyclical trend. 


### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}
nobs <- nrow(data) 
#Create vector t
t <- c(1:nobs)

#Fit a linear trend to TS of Total Biomass Energy Production
linear_trend_model1=lm(data[,2]~t)  
print(summary(linear_trend_model1))
beta1.0=as.numeric(linear_trend_model1$coefficients[1])  #first coefficient is the intercept term or beta0
beta1.1=as.numeric(linear_trend_model1$coefficients[2])  #second coefficient is the slope or beta1
  
#Let's plot the time series with its trend line
print(ggplot(data, aes(x=Date, y=data[,(2)])) +
        geom_line(color="blue") +
        ylab(paste0(colnames(data)[2], " (Trillion Btu)", sep = " "))+
        xlab("Year")+
        labs(title = paste0(colnames(data)[2], " Over Time", sep = " "))+
        #geom_abline(intercept = beta0, slope = beta1, color="red")
        geom_smooth(color="red",method="lm"))
```
There is a significantly positive correlation between the Total Biomass Energy Production and time. An intercept of 133.7 indicates that in year 1973, there was about 133.7 trillion Btu of Total Biomass Energy Produced. The slope of 0.48 indicates that every year, the Total Biomass Energy Produced increases by 0.40 trillion Btu.  

```{r}
#Fit a linear trend to TS of Total Renewable Energy Production
linear_trend_model2=lm(data[,3]~t)  
print(summary(linear_trend_model2))
beta2.0=as.numeric(linear_trend_model2$coefficients[1])  #first coefficient is the intercept term or beta0
beta2.1=as.numeric(linear_trend_model2$coefficients[2])  #second coefficient is the slope or beta1

#Let's plot the time series with its trend line
print(ggplot(data, aes(x=Date, y=data[,(3)])) +
        geom_line(color="blue") +
        ylab(paste0(colnames(data)[3], " (Trillion Btu)", sep = " "))+
        xlab("Year")+
        labs(title = paste0(colnames(data)[3], " Over Time", sep = " "))+
        #geom_abline(intercept = beta0, slope = beta1, color="red")
        geom_smooth(color="red",method="lm"))
```
There is also a significantly positive correlation between Total Renewable Energy Production and time. An intercept of 312.2 indicates that in 1973, there was about 312.2 trillion Btu of Total Renewable Energy Produced. The slope of 0.9362 indicates that every year, the Total Renewable Energy Produced increases by 0.9362 trillion Btu. 

```{r}
#Fit a linear trend to TS of Hydroelectric Power Consumption
linear_trend_model3=lm(data[,4]~t)  
print(summary(linear_trend_model3))
beta3.0=as.numeric(linear_trend_model3$coefficients[1])  #first coefficient is the intercept term or beta0
beta3.1=as.numeric(linear_trend_model3$coefficients[2])  #second coefficient is the slope or beta1
  
#Let's plot the time series with its trend line
print(ggplot(data, aes(x=Date, y=data[,(4)])) +
        geom_line(color="blue") +
        ylab(paste0(colnames(data)[4], " (Trillion Btu)", sep = " "))+
        xlab("Year")+
        labs(title = paste0(colnames(data)[4], " Over Time", sep = " "))+
        #geom_abline(intercept = beta0, slope = beta1, color="red")
        geom_smooth(color="red",method="lm"))

```
There is a significantly negative trend between Hydroelectric Power Consumption and time. In 1973, the Hydroelectric Power Consumption was at about 259.9 trillion Btu. Every year, Hydroelectric Power Consumption decreases by 0.083 trillion Btu. 
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}
i<- 1
#remove the trend from series
detrend_data1 <- data[,(i+1)]-(beta1.0+beta1.1*t)
#plot graph
ggplot(data, aes(x=Date, y=data[,(1+i)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[i+1], " (Trillion Btu)", sep = " "))+
  xlab("Year")+
  #geom_abline(intercept = beta0, slope = beta1, color="red")
  geom_smooth(color="red",method="lm") +
  geom_line(aes(y=detrend_data1), col="green")+
  geom_smooth(aes(y=detrend_data1),color="orange",method="lm")+
  labs(title = paste0(colnames(data)[i+1], " Over Time (in blue) and Detrended (in green)", sep = " "))
```
The detrended series no longer has a positive correlation with time. The mean is now at zero. 

```{r}
i<- 2
#remove the trend from series
detrend_data2 <- data[,(i+1)]-(beta2.0+beta2.1*t)
#plot graph
ggplot(data, aes(x=Date, y=data[,(1+i)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[i+1], " (Trillion Btu)", sep = " "))+
  xlab("Year")+
  #geom_abline(intercept = beta0, slope = beta1, color="red")
  geom_smooth(color="red",method="lm") +
  geom_line(aes(y=detrend_data2), col="green")+
  geom_smooth(aes(y=detrend_data2),color="orange",method="lm")+
  labs(title = paste0(colnames(data)[i+1], " Over Time (in blue) and Detrended (in green)", sep = " "))
```
The detrended series no longer has a positive correlation with time.

```{r}
i<- 3
#remove the trend from series
detrend_data3 <- data[,(i+1)]-(beta3.0+beta3.1*t)
#plot graph
ggplot(data, aes(x=Date, y=data[,(1+i)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[i+1], " (Trillion Btu)", sep = " "))+
  xlab("Year")+
  #geom_abline(intercept = beta0, slope = beta1, color="red")
  geom_smooth(color="red",method="lm") +
  geom_line(aes(y=detrend_data3), col="green")+
  geom_smooth(aes(y=detrend_data3),color="orange",method="lm")+
  labs(title = paste0(colnames(data)[i+1], " Over Time (in blue) and Detrended (in green)", sep = " "))
```
The detrended series no longer has a negative correlation with time. It is easier to see the cyclicity within the time series now.

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

```{r}
#make data frame of detrended data
detrend_data<- data.frame(detrend_data1)
detrend_data$i2<- detrend_data2
detrend_data$i3<- detrend_data3
detrend_data$Date<- data$Date

detrend_data<- detrend_data[,c(4,1:3)]
colnames(detrend_data)<- colnames(data)
ts_detrend_data<- ts(detrend_data[,(2:4)], frequency = 12, start = c(1973,1))

#plot time series, Acf, and Pacf
for(i in 1:(ncol(detrend_data)-1)){
  Acf(detrend_data[,i], lag.max=40, main=name[i+1], ylim=c(-1,1))
  Pacf(detrend_data[,i], lag.max=40,main=name[i+1], ylim=c(-1,1))
}
```
After detrending, there is no longer a strong correlation between any lags in the Total Biomass Energy Production time series. We continue to see the seasonality in the ACF plot of the Hydroelectric Power Consumption time series. 

## Seasonal Component
Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

```{r}
#Total Biomass Energy Production
j=1 

#create the seasonal dummies
dummies1 <- seasonaldummy(ts_data[,j])  
#this function only accepts ts object, no need to add one here because date object is not a column

#fit a linear model to the seasonal dummies
seas_means_model1=lm(data[,(j+1)]~dummies1)
summary(seas_means_model1)

#store regression coefficients
beta_int1=seas_means_model1$coefficients[1]
beta_coeff1=seas_means_model1$coefficients[2:12]

#compute seasonal component
inflow_seas_comp1=array(0,nobs)
for(i in 1:nobs){
  inflow_seas_comp1[i]=(beta_int1+beta_coeff1%*%dummies1[i,])
}

#graph 
ggplot(data, aes(x=Date, y=data[,(j+1)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[j+1], " (Trillion Btu)", sep = " "))+
  geom_line(aes(y=inflow_seas_comp1), col="red")
```
This series does not have a seasonal trend. The time series is not significantly correlated with any of the dummy months. 

```{r}
#Total Renewable Energy Production
j=2

#create the seasonal dummies
dummies2 <- seasonaldummy(ts_data[,j])  
#this function only accepts ts object, no need to add one here because date object is not a column

#fit a linear model to the seasonal dummies
seas_means_model2=lm(data[,(j+1)]~dummies2)
summary(seas_means_model2)

#store regression coefficients
beta_int2=seas_means_model2$coefficients[1]
beta_coeff2=seas_means_model2$coefficients[2:12]

#compute seasonal component
inflow_seas_comp2=array(0,nobs)
for(i in 1:nobs){
  inflow_seas_comp2[i]=(beta_int2+beta_coeff2%*%dummies2[i,])
}

#graph 
ggplot(data, aes(x=Date, y=data[,(j+1)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[j+1], " (Trillion Btu)", sep = " "))+
  geom_line(aes(y=inflow_seas_comp2), col="red")
```
This series does not have a seasonal trend. The time series is not significantly correlated with any of the dummy months. 

```{r}
#Hydroelectric Power Consumption
j=3

#create the seasonal dummies
dummies3 <- seasonaldummy(ts_data[,j])  
#this function only accepts ts object, no need to add one here because date object is not a column

#fit a linear model to the seasonal dummies
seas_means_model3=lm(data[,(j+1)]~dummies3)
summary(seas_means_model3)

#store regression coefficients
beta_int3=seas_means_model3$coefficients[1]
beta_coeff3=seas_means_model3$coefficients[2:12]

#compute seasonal component
inflow_seas_comp3=array(0,nobs)
for(i in 1:nobs){
  inflow_seas_comp3[i]=(beta_int3+beta_coeff3%*%dummies3[i,])
}

#graph 
ggplot(data, aes(x=Date, y=data[,(j+1)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[j+1], " (Trillion Btu)", sep = " "))+
  geom_line(aes(y=inflow_seas_comp3), col="red")
```
This series has a strong seasonal component (p-value < 0.05). The time series is significantly correlated with all but two of the dummy months. 

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?
```{r}
#Total Biomass Energy Production
j<-1
#Removing seasonal component
deseason_data1 <- data[,(1+j)]-inflow_seas_comp1

#graph
ggplot(data, aes(x=Date, y=data[,(j+1)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[j+1], " (Trillion Btu)", sep = " "))+
  geom_line(aes(y=deseason_data1), col="green")+
  labs(title = paste0(colnames(data)[j+1], " Over Time (in blue) and Deseasoned (in green)", sep = " "))
```
There is virtually no change in the shape of the data, only in the vertical position of the data, because there is little to no seasonality in the data. 

```{r}
#Total Renewable Energy Production
j<-2
#Removing seasonal component
deseason_data2 <- data[,(1+j)]-inflow_seas_comp2

#graph
ggplot(data, aes(x=Date, y=data[,(j+1)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[j+1], " (Trillion Btu)", sep = " "))+
  geom_line(aes(y=deseason_data2), col="green")+
  labs(title = paste0(colnames(data)[j+1], " Over Time (in blue) and Deseasoned (in green)", sep = " "))
```
There is little change in the shape of the data due to very little seasonality in the data. 

```{r}
#Hydroelectric Power Consumption
j<-3
#Removing seasonal component
deseason_data3 <- data[,(j+1)]-inflow_seas_comp3

#graph
ggplot(data, aes(x=Date, y=data[,(j+1)])) +
  geom_line(color="blue") +
  ylab(paste0(colnames(data)[j+1], " (Trillion Btu)", sep = " "))+
  geom_line(aes(y=deseason_data3), col="green")+
  labs(title = paste0(colnames(data)[j+1], " Over Time (in blue) and Deseasoned (in green)", sep = " "))
```
In the deseasoned graph, we can see the differences between each point a lot better than in the original data. 

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?
```{r}
#make data frame of deseasoned data
deseason_data<- data.frame(deseason_data1)
deseason_data$i2<- deseason_data2
deseason_data$i3<- deseason_data3
deseason_data$Date<- data$Date

deseason_data<- deseason_data[,c(4,1:3)]
colnames(deseason_data)<- colnames(data)
ts_deseason_data<- ts(deseason_data[,(2:4)], frequency = 12, start = c(1973,1))

#plot time series, Acf, and Pacf
for(i in 1:(ncol(deseason_data)-1)){
  Acf(ts_deseason_data[,i], lag.max=40, main=name[i+1], ylim=c(-1,1))
  Pacf(ts_deseason_data[,i], lag.max=40,main=name[i+1], ylim=c(-1,1))
}
```
After removing the seasonality from the Total Renewable Energy Production and Hydroelectric Power Consumption time series, there is not longer any autocorrelation between any lags in the PACF plot. Acording to the ACF plot of the Hydroelectric Power Consumption time series, it is autoregressive. 
H
