---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 5 - Due date 02/27/23"
author: "Angela Zeng"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A05_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(readxl)
library(forecast)
library(tseries)
library(ggplot2)
library(Kendall)
library(lubridate)
library(tidyverse)  #load this package so you clean the data frame using pipes
```

## Decomposing Time Series

Consider the same data you used for A04 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx".

```{r}
#Importing data set - using xlsx package
energy_data<- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE)

#Now let's extract the column names from row 11 only
read_col_names <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(energy_data) <- read_col_names
head(energy_data)

nobs=nrow(energy_data)
nvar=ncol(energy_data)
```

### Q1

For this assignment you will work only with the following columns: Solar Energy Consumption and Wind Energy Consumption. Create a data frame structure with these two time series only and the Date column. Drop the rows with *Not Available* and convert the columns to numeric. You can use filtering to eliminate the initial rows or convert to numeric and then use the drop_na() function. If you are familiar with pipes for data wrangling, try using it!

```{r}
#transform Date column 
energy_data$Month<- as.Date(energy_data$Month)
colnames(energy_data)[1]<- "Date"
head(energy_data)

#extract columns 
colnames(energy_data)
data<- energy_data[,c(1,8,9)]
data<- data %>%
  filter(`Solar Energy Consumption` != "Not Available")
data$`Solar Energy Consumption`<- as.numeric(data$`Solar Energy Consumption`)
data$`Wind Energy Consumption`<- as.numeric(data$`Wind Energy Consumption`)

head(data)
tail(data) #no NAs
```

### Q2

Plot the Solar and Wind energy consumption over time using ggplot. Plot each series on a separate graph. No need to add legend. Add informative names to the y axis using `ylab()`. Explore the function scale_x\_date() on ggplot and see if you can change the x axis to improve your plot. Hint: use *scale_x\_date(date_breaks = "5 years", date_labels = "%Y")")*

```{r}
ggplot(data, aes(x=Date , y=`Solar Energy Consumption`)) +
  geom_line(color= "red") + 
  ylab('Solar Energy Consumption (Trillion Btu)')+
  scale_x_date(date_breaks = "5 years", date_labels = "%Y")

ggplot(data, aes(x=Date , y=`Wind Energy Consumption`)) +
  geom_line(color= "blue") + 
  ylab('Wind Energy Consumption (Trillion Btu)')+
  scale_x_date(date_breaks = "5 years", date_labels = "%Y")
```

### Q3

Now plot both series in the same graph, also using ggplot(). Look at lines 141-148 of the file `M4_OutliersMissingData_Part2_Complete.Rmd` to learn how to manually add a legend to ggplot. Make the solar energy consumption red and wind energy consumption blue. Add informative name to the y axis using `ylab("Energy Consumption)`. And use function scale_x\_date() again to improve x axis.

```{r}
ggplot(data) +
  geom_line(aes(x = Date, y = `Solar Energy Consumption`, color = "Solar")) +
  geom_line(aes(x = Date, y = `Wind Energy Consumption`, color = "Wind")) +
  scale_color_manual(values = c("Solar" = "red", "Wind" = "blue"))+
  theme(legend.position = "bottom") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y")+
  labs(color="",
       title = 'Energy Consumption Over Time',
       y='Energy Consumption (Trillion Btu)',
       x='Year')
```

### Q3

Transform wind and solar series into a time series object and apply the decompose function on them using the additive option, i.e., `decompose(ts_data, type = "additive")`. What can you say about the trend component? What about the random component? Does the random component look random? Or does it appear to still have some seasonality on it?

```{r}
#make time series data 
ts_solar<- ts(data[,2], frequency = 12, start = c(1984,1))
ts_wind<- ts(data[,3], frequency = 12, start = c(1984,1))

decompose_solar<- decompose(ts_solar, type = "additive")
plot(decompose_solar) 
title(sub = "Solar Energy Consumption", col.sub = "blue")

decompose_wind<- decompose(ts_wind, type = "additive")
plot(decompose_wind)
title(sub = "Wind Energy Consumption", col.sub = "blue")
```
>Solar energy consumption looks like it has an increasing exponential trend. The random component of solar energy consumption is not random; there appears to be some seasonality in it. Starting a few years before 2010, wind energy consumption has an increasing linear trend. The random component of wind energy consumption looks like it could be random. 

### Q4

Use the decompose function again but now change the type of the seasonal component from additive to multiplicative. What happened to the random component this time?

```{r}
decompose_solar_m<- decompose(ts_solar, type = "multiplicative")
plot(decompose_solar_m)
title(sub = "Solar Energy Consumption", col.sub = "blue")

decompose_wind_m<- decompose(ts_wind, type = "multiplicative")
plot(decompose_wind_m)
title(sub = "Wind Energy Consumption", col.sub = "blue")
```
>When using the multiplicative seasonal component, the random component flips to having more randomness during the earlier years of the time series, rather than the later years of the series. 

### Q5

When fitting a model to this data, do you think you need all the historical data? Think about the data from 90s and early 20s. Are there any information from those years we might need to forecast the next six months of Solar and/or Wind consumption. Explain your response.

> Answer: When fitting a model to this data, we probably don't need data from when there was no or very little solar and wind energy consumption. The magnitude of solar and wind consumption is now so great (in the hundreds of Trillions Btu), that 0.01 Trillion Btu of solar energy consumption in 1984 will not be helpful in determining solar energy consumption in the future. 

### Q6

Create a new time series object where historical data starts on January 2012. Hint: use `filter()` function so that you don't need to point to row numbers, .i.e, `filter(xxxx, year(Date) >= 2012 )`. Apply the decompose function `type=additive` to this new time series. Comment the results. Does the random component look random? Think about our discussion in class about seasonal components that depends on the level of the series. 

```{r}
data_2012<- data %>%
  filter(year(Date) >= 2012)

#make time series data 
ts_solar12<- ts(data_2012[,2], frequency = 12, start = c(2012,1))
ts_wind12<- ts(data_2012[,3], frequency = 12, start = c(2012,1))

#decompose 
decompose_solar12<- decompose(ts_solar12, type = "additive")
plot(decompose_solar12) 
title(sub = "Solar Energy Consumption Starting in 2012", col.sub = "blue")
    
decompose_wind12<- decompose(ts_wind12, type = "additive")
plot(decompose_wind12)
title(sub = "Wind Energy Consumption Starting in 2012", col.sub = "blue")
```

> Answer: The random component of the solar energy consumption, starting in 2012, still does not look random. There seems to be a seasonal component still. However, the random component of wind energy consumption does look random. 



